Higher order unification for free!

### 2024-04-08 paper/main.tex
move related works to the end
major reorganization of 2,3,4... still a wip

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern-fragment.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}

To state precisely the problem we solve we need a \Fo{} representation
of CIC terms and a \Ho one. We extend the equality over ground
terms to open ones with the reflexivity of unification variables (that is
unification variable X is equal to itself, not to another one such a Y).
We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
We call \Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
We assume that the unification of our meta
language is correct and complete in \llambda.
%
%\begin{gather}
%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
%\end{gather}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$ and we use set union to denote the concatenation of
two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
$\sigma$ for the \Ho ones.

@@

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$.
Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
%$\foUnifPb = \{ s_1, s_2 \}$ 
as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

\begin{proposition}[Properties of \Uo{} in \llambda]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
\end{gather}
\end{proposition}

\noindent
In addition to correctness and completeness in \llambda, we claim that \hrun
handles terms outside \llambda in the following sense:

@@

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory \emph{link} that represent a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
contradict \Eo as it would otherwise do on ``problematic'' terms.
We now define ``problematic'' and ``well behaved'' more formally.

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
since decompilation can introduce (actually restore) terms in
\maybeeta, \maybebeta or \notllambda that were moved out of the way
during compilation.

\section{Preliminaries: \Fo{} and \Ho}
\label{sec:lang-spec}

\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term \coqIn{P x} is represented as
\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
since unification variables come equipped with an explicit scope.
We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

@@

@@

\subsection{Equational theory and Unification}

In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
we need to equip \Fo{} and \Ho with term equality,
substitution application and unification.

\input{code/ue_type}

@@

\subsubsection{Detection of \maybebeta}

>>>>

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories and
unification procedures of the meta language and the object language.

The equational theory of CIC is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:\\
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

Even if we encountered the problem in CIC one does not need such a
complex system in order to state it. The minimal setting we could find
is one where a \Fo{} language (for first order) with a rich(er) equational
theory is phrased in a \Ho{} meta language with a simpler one, and one
needs to build the unification of \Fo{} on top of the one of \Ho{}.

\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term \coqIn{P x} is represented as
\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
since unification variables come equipped with an explicit scope.
\todo{use faddr for Fo}

faddr and addr are memory addresses, the details are given in bla,
here they are just unique identifiers for unif variables and their types
do now allow to mistake one for the other.
\todo{cleanup}

\subsection{eq and unif}

In order to specify the two unification procedures we need to
define the equational theories of \Fo{} and \Ho{} and define
what substitutions are.

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}
that we shall define later.\todo{fixup}

\input{code/ue_type}
%
% To state precisely the problem we solve we need a \Fo{} representation
% of CIC terms and a \Ho one. We extend the equality over ground
% terms to open ones with the reflexivity of unification variables (that is
% unification variable X is equal to itself, not to another one such a Y).
% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
% We call \Uo the unification procedure we want to implement and
% \Ue the one provided by the meta language.
We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$ and we use set union to denote the concatenation of
two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
$\sigma$ for the \Ho ones.
We assume that the unification of our meta
language is correct and complete in \llambda.

\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

\subsection{Logic program simulation}

@@

@@

\noindent 
\todo{later in section..}
We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$.
Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
% %$\foUnifPb = \{ s_1, s_2 \}$ 
% as follows:
% %
% $$
% \begin{array}{l}
% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
% \end{array}
% $$

% \begin{proposition}[Properties of \Uo{} in \llambda]
% \begin{gather}
%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
% \end{gather}
% \end{proposition}

\noindent
%In addition to correctness and completeness in \llambda,
We claim that \hrun
handles terms outside \llambda in the following sense:

@@

\subsection{The solution in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory \emph{link} that represent a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on ``problematic'' terms.
We now define ``problematic'' and ``well behaved'' more formally.

@@

% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) terms in
% \maybeeta, \maybebeta or \notllambda that were moved out of the way
% during compilation.

\section{Preliminaries for \Fo{} and \Ho}
\label{sec:lang-spec}
\todo{maybe move here napp beta deref fderef move}

We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

@@

%\input{code/move}

@@

\todo{notation for problem compilation}

@@

\subsubsection{Compilation of terms in \maybebeta}
